---
title: "2 Tipos e estruturas de dados em R"
output: html_notebook
author: Prof. Eduardo 
---


# Tipos de Dados (*Classes e Objetos*)
__Tudo em R é objeto!!__
R tem cinco classes básicas ou *atômicas* de objetos:
-   caractere (string)
-   numérico (números reais)
-   inteiro
-   números complexos 
-   Valores lógicos (TRUE/FALSE)

```{r}
#Caractere
x <- "Bem vindo"
```

## Números
-   Números em R são geralmente tratados como objetos (ex. double
    precision real numbers)
-   Se você precisa especificamente de um inteiro (raramente), é necessário especificar o sufuxo ```L```

-   O valor ```NaN``` representa um valor não identificado (*"not a number"*);
    e.g. 0 / 0; ```NaN``` pode também representar valores ausentes (não se preocupe agora com isso, mas mostra que a linguagem foi concebida para identificar essa situação )

```{r}
#Numérico
x <- 5
x
is.numeric(x)
is.integer(x)
```
```{r}
#Inteiro (apenas um detalhe. Raramente é importante)
x <- 5L
x
is.numeric(x)
is.integer(x)
```
```{r}
#Números Complexos (Uso não muito comum)
x <- 1 + 4i
x
is.numeric(x)
is.integer(x)
is.complex(x)
```
```{r}
#Lógico
x <- TRUE
x
x <- True

x<-(3 < 4)
x
```

#Estruturas de dados

##Vetores
A estrutura mais básica do __R__ é um vetor
-   Um vetor pode conter apenas objetos da mesma classe básica
-   Existe um outro objeto chamado *lista*, que é representado como um vetor, mas pode conter objetos de classes diferentes
-   Vetores vazios podem ser criados com a função ```vector() ```.
- O operador `:` é usado para gerar/representar uma sequência de números.

```r
 x <- 1:20 
 x
```

## Criando Vetores

A função `c()` concatena os seus argumentos em um vetor.

```r
 x <- c(0.5, 0.6)       ## numeric
 
 x <- c(TRUE, FALSE)    ## logical
 
 x <- c(T, F)           ## logical
 
 x <- c("a", "b", "c")  ## character
 
 x <- 9:29              ## integer
 
 x <- c(1+0i, 2+4i)     ## complex
 
```

Para a criação de um vetor *vazio* podemos usar a função `vector()`

```r
 x <- vector("numeric", length = 10) 
 x
 [1] 0 0 0 0 0 0 0 0 0 0
```


## Attributes

Os objetos em possuem vários atributos:

    -   nomes, dimnames
    -   dimensão (e.g. matrizes, arrays)
    -   classe
    -   comprimento
    -   outras atributos definidos pelo usuário/metadata

O nome dos atributos de um objeto pode ser acessado usando  a função `attributes()`.


```r
attributes(x)"
```


## Misturando Objetos

Quando objetos diferente são misturados em um vetor,  a *coerção* ajusta os tipos de modo que todos elementos mantenham sejam da mesma classe.

```r
 y <- c(1.7, "a")   ## character
 y <- c(TRUE, 2)    ## numeric
 y <- c("a", TRUE)  ## character
```



## Coerção explícita

Objetos de uma classe podem ser sua classe alterada usando as funções `as.*()`

```r
 x <- 0:6
 class(x)          # "integer"
 as.numeric(x)  
 as.logical(x)     # FALSE ....
 as.character(x)   # "0" "1" "2" "3" "4" "5" "6"
```

Coerções não permitidas geram  `NA` (Não avaliado).

```r
 x <- c("a", "b", "c")
 as.numeric(x)
 as.logical(x)
 as.complex(x)
```


## Matrices

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Matrices are vectors with a _dimension_ attribute. The dimension attribute is itself an integer vector of length 2 (nrow, ncol)

```r
 m <- matrix(nrow = 2, ncol = 3) 
 m
     [,1] [,2] [,3]
[1,]   NA   NA   NA
[2,]   NA   NA   NA
 dim(m)
[1] 2 3
 attributes(m)
$dim
[1] 2 3
```

---

## Matrices (cont'd)

Matrices are constructed _column-wise_, so entries can be thought of starting in the "upper left" corner and running down the columns.

```r
 m <- matrix(1:6, nrow = 2, ncol = 3) 
 m
     [,1] [,2] [,3]
[1,]    1    3    5
[2,]    2    4    6
```

---

## Matrices (cont'd)

Matrices can also be created directly from vectors by adding a dimension attribute.

```r
 m <- 1:10 
 m
[1] 1 2 3 4 5 6 7 8 9 10 
 dim(m) <- c(2, 5)
 m
     [,1] [,2] [,3] [,4] [,5]
[1,]    1    3    5    7    9
[2,]    2    4    6    8   10
```

---

## cbind-ing and rbind-ing

Matrices can be created by _column-binding_ or _row-binding_ with `cbind()` and `rbind()`.

```r
 x <- 1:3
 y <- 10:12
 cbind(x, y)
     x  y 
[1,] 1 10 
[2,] 2 11 
[3,] 3 12
 rbind(x, y) 
  [,1] [,2] [,3]
x    1    2    3
y   10   11   12
```

---

## Lists

Lists are a special type of vector that can contain elements of different classes. Lists are a very important data type in R and you should get to know them well.

```r
 x <- list(1, "a", TRUE, 1 + 4i) 
 x
[[1]]
[1] 1

[[2]] 
[1] "a"

[[3]]
[1] TRUE

[[4]]
[1] 1+4i
```

---

## Factors

Factors are used to represent categorical data. Factors can be unordered or ordered. One can think of a factor as an integer vector where each integer has a _label_.

- Factors are treated specially by modelling functions like `lm()` and `glm()`

- Using factors with labels is _better_ than using integers because factors are self-describing; having a variable that has values "Male" and "Female" is better than a variable that has values 1 and 2.

---

## Factors

```r
 x <- factor(c("yes", "yes", "no", "yes", "no")) 
 x
[1] yes yes no yes no
Levels: no yes
 table(x) 
x
no yes 
 2   3
 unclass(x)
[1] 2 2 1 2 1
attr(,"levels")
[1] "no"  "yes"
```

---

## Factors

The order of the levels can be set using the `levels` argument to `factor()`. This can be important in linear modelling because the first level is used as the baseline level.

```r
 x <- factor(c("yes", "yes", "no", "yes", "no"),
              levels = c("yes", "no"))
 x
[1] yes yes no yes no 
Levels: yes no
```

---

## Missing Values

Missing values are denoted by `NA` or `NaN` for undefined mathematical operations. 

- `is.na()` is used to test objects if they are `NA`

- `is.nan()` is used to test for `NaN`

- `NA` values have a class also, so there are integer `NA`, character `NA`, etc.

- A `NaN` value is also `NA` but the converse is not true

---

## Missing Values

```r
 x <- c(1, 2, NA, 10, 3)
 is.na(x)
[1] FALSE FALSE  TRUE FALSE FALSE
 is.nan(x)
[1] FALSE FALSE FALSE FALSE FALSE
 x <- c(1, 2, NaN, NA, 4)
 is.na(x)
[1] FALSE FALSE  TRUE  TRUE FALSE
 is.nan(x)
[1] FALSE FALSE  TRUE FALSE FALSE
```

---

## Data Frames

Data frames are used to store tabular data

- They are represented as a special type of list where every element of the list has to have the same length

- Each element of the list can be thought of as a column and the length of each element of the list is the number of rows

- Unlike matrices, data frames can store different classes of objects in each column (just like lists); matrices must have every element be the same class

- Data frames also have a special attribute called `row.names`

- Data frames are usually created by calling `read.table()` or `read.csv()`

- Can be converted to a matrix by calling `data.matrix()`

---

## Data Frames

```r
 x <- data.frame(foo = 1:4, bar = c(T, T, F, F)) 
 x
  foo   bar
1   1  TRUE
2   2  TRUE
3   3 FALSE
4   4 FALSE
 nrow(x)
[1] 4
 ncol(x)
[1] 2
```

---

## Names

R objects can also have names, which is very useful for writing readable code and self-describing objects.

```r
 x <- 1:3
 names(x)
NULL
 names(x) <- c("foo", "bar", "norf") 
 x
foo bar norf 
  1   2    3
 names(x)
[1] "foo"  "bar"  "norf"
```

---

## Names

Lists can also have names.

```r
 x <- list(a = 1, b = 2, c = 3) 
 x
$a
[1] 1

$b 
[1] 2

$c 
[1] 3
```

---

## Names

And matrices.

```r
 m <- matrix(1:4, nrow = 2, ncol = 2)
 dimnames(m) <- list(c("a", "b"), c("c", "d")) 
 m
  c d 
a 1 3 
b 2 4
```

---

## Summary

Data Types

- atomic classes: numeric, logical, character, integer, complex \

- vectors, lists

- factors

- missing values

- data frames

- names
